The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-10-06 thesis/acronyms.tex
thesis updates / reorganization of RV modeling
major thesis restructure
julio edits at durham
thesis julio ch1-3 edits
thesis updates
thesis; julio oct 4 comments

% Ordinary abbreviations
\newcommand{\kms}{{\rm km\,s^{-1}}}
\newcommand{\kmsdeg}{{\rm km\,s^{-1}\,deg^{-1}}}

@@

\newacronym
[description={
Apache Point Observatory Galactic Evolution Experiment. 
A large near-infrared spectroscopic survey of stars in the Milky Way. \citep{apogee17.}}]
{apogee}{apogee}{Apache Point Observatory Galactic Evolution Experiment}

>>>>

% Ordinary abbreviations
% units
\newcommand{\kms}{{\rm km\,s^{-1}}}
\newcommand{\kmsdeg}{{\rm km\,s^{-1}\,deg^{-1}}}
\newcommand{\masyr}{{\rm mas^{-1}\,yr^{-1}}}
\newcommand{\Gyr}{{\rm Gyr}}
\newcommand{\kpc}{{\rm kpc}}
\newcommand{\Msun}{{\rm M_\odot}}

% match shortcuts
\newcommand{\pmra}{\mu_{\alpha*}}
\newcommand{\pmdec}{\mu_{\delta}}
\newcommand{\V}{{\rm v}}
\newcommand{\vmax}{\V_{\rm max}}
\newcommand{\vcirc}{\V_{\rm circ}}
\newcommand{\rmax}{r_{\rm max}}

\newcommand{\LCDM}{$\Lambda$CDM}
\newcommand{\agama}{{\sc Agama}}
\newcommand{\gadget}{{\sc gadget-4}}
\newcommand{\apostle}{{\sc apostle}}
\newcommand{\eagle}{{\sc eagle}}
\newcommand{\smallperi}{{\tt smallperi}}

@@

{apogee}{apogee}{Apache Point Observatory Galactic Evolution Experiment}

### 2025-10-06 thesis/data.tex
thesis; julio oct 4 comments

\section{Density profiles}\label{density-profiles}

@@

Our density profiles are robust to alternative methodologies and
magnitude cuts. Density profiles based on J+24 candidates may be biased
by their assumed spatial likelihoods. To address this, we consider a
Bayesian model with a non-parametric spatial likelihood. We also explore
selections using absolute cuts, and the effect of structural assumptions
of the density profile (Appendix \ref{sec:extra_density}). These
methodologies are consistent except when the derived density drops below
the background of satellite-like stars (from the CMD+PM sample). We also
do not find evidence of magnitude biases in the density profiles,
indicating inhomogeneous magnitude-completeness is small. Our density
profiles furthermore agree with photometric surveys and similar
literature (see Section~\ref{sec:extra_density}). We conclude that
selection criteria do not influence our results.

>>>>

\section{Density profiles}\label{sec:data_density_profiles}

@@

Our density profiles are robust to alternative methodologies and
magnitude cuts. Density profiles based on J+24 candidates may be biased
by their assumed spatial likelihoods. To address this, we consider a
Bayesian model with a non-parametric spatial likelihood. We also explore
selections using absolute cuts, and the effect of structural assumptions
of the density profile (Appendix \ref{sec:extra_density}). These
methodologies are consistent except when the derived density drops below
the background of satellite-like stars (from the CMD+PM sample). We also
do not find evidence of magnitude biases in the density profiles,
indicating inhomogeneous magnitude-completeness is small. Our density
profiles furthermore agree with photometric surveys and similar
literature (see Appendix \ref{sec:extra_density}). We conclude that
selection criteria do not influence our results.

### 2025-10-06 thesis/discussion.tex
thesis; julio oct 4 comments

In the previous chapters, we have shown that Sculptor and Ursa Minor
have more extended density profiles compared to other MW classical
satellites, and that recent tides are an unlikely explanation for such
extended density profiles. We therefore suggest that Scl and UMi host a
``stellar halo''---an additional, extended population of stars. With our
data, other classical dwarfs lack evidence for similar stellar halos.
The absence of similar halos in other classical dwarfs suggests Scl and
UMi have followed a divergent evolutionary path.

In this chapter, we place our results in context with past work. We
review both work considering tidal effects on Scl and UMi and detections
of peculiarities. We then consider possible origins of stellar halos for
these galaxies. Scenarios include induced star formation forming
multiple populations, past mergers depositing stars into the outskirts,
and even departures from the standard dwarf galaxy structure. We later
discuss how future observations and simulations may distinguish halo
formation scenarios. Finally, we summarize the main conclusions of this
thesis.

@@

Tidal effects have been long invoked to explain features of dwarf
galaxies e.g. \ref{sec:scl_umi_obs_tides},
\citep{tsujimoto+shigeyama2002, mayer+2001a}. Yet, growing work suggests
tides may not be as commonplace as suspected. For example,
\citet{read+2006} suggest that the absence of rising radial velocity
dispersion profiles in dwarfs implies a lack of tidal disruption.
\citet{penarrubia+2009}, introducing the break radius as a diagnostic
(Eq.~\ref{eq:r_break}), showing that most satellite orbits were
inconsistent with observable tidal effects. Recently,
\citet{pace+erkal+li2022} use a Jacobi radius-based criterion
(Eq.~\ref{eq:r_jacobi}), reaching a similar conclusion. With the
exception of the Sagittarius dSph, these results suggest tides have mild
influence on the stellar components of classical dwarfs.

@@

Our work extends these models, adding updated structural properties, a
broader range of orbital histories, and the influence of the LMC.
Despite these considerations, we still reach similar conclusions: tidal
effects do not shape Scl or UMi's stellar component. A non-tidal
explanation likely underlies the origin of these galaxies' density
excess.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_umi_fe_h_gradient.pdf}}
\caption[Metallicity gradients in Scl and UMi]{Density profiles and
metallicity gradients in Scl (left) and UMi (right). In the top panel,
we plot the J+24 density profiles as function of elliptical radius in
black, and the two components of a double-exponential density fit as
curved, transparent lines. The vertical lines represent the transition
from the inner to the outer exponential profile. \textbf{Bottom}:
Metallicies as a function of elliptical radius for stars from
\citet{tolstoy+2023}; \citet{apogee}; and \citet{sestito+2023} (left)
and \citet{pace+2020}; \citet{apogee}; and \citet{sestito+2023b}
(right)}\label{fig:metallicity_gradients}
\end{figure}

@@

Could these internal stellar populations map to the observed density
excess? In Scl, both populations have scale radii likely smaller than
the half-light radius. But, the third, more tentatively detected
population in \citet{arroyo-polonio+2024} would reside within the
stellar halo. In Ursa Minor, the transition between the metal-rich and
metal-poor stellar components occurs at \(\sim 30\,\)arcmin
\citep{pace2020}, consummate with the beginning of the density excess.
This metal-poor population may thus be related to the stellar halo of
Ursa Minor. Fig:metallicity\_gradients shows the radial metallicity
distribution in both galaxies. Scl and UMi both show evidence of a
metallicity gradient, but the outskirts have few stars. Further
observations would be necessary to understand if the stellar halo
represents a chemically distinct population.

Ursa Minor has also shown evidence for possible inner substructure such
as stellar or kinematic ``clumps''
\citep[e.g.,][]{olszewski+aaronson1985, demers+1995, kleyna+1998, battinelli+demers1999, bellazzini+2002}.
While one clump was re-detected kinematically in \citet{pace+2014},
\citet{munoz+2018} find no evidence of substructure with modern
photometry. In any case, the survival of a cold clump depends on the
nature of dark matter, disrupting quickly in a \LCDM{} halo
\citep{kleyna+2003, lora+2012}.

Given their multiple stellar populations and possible inner
substructure, this hints that Scl and UMi's distinct structure today
arises from an unusual past. A mechanism forming a stellar halo could
also explain multiple populations.

\section{Forming a stellar halo}\label{sec:stellar_halos}

As we disfavour a tidal origin for Scl and UMi's extended density
profiles, we now consider alternative processes which may form a stellar
halo in these dwarfs.

\textbf{Episodic star formation.} Star formation may quench and
reignite, creating successive stellar generations with differing
distributions. External star formation triggers include tidal
compression \citep{mayer+2001a, dong+lin+murray2003}, collisions with
gaseous filaments \citep{genina+2019}, perturbations from dark halos
\citep{starkenburg+helmi+sales2016}, or shocks with the MW corona
\citep{wright+2019}. More common mechanisms, like feedback or
reionization-driven quenching, may also form multiple stellar
generations
\citep{kawata+2006, benitez-llambay+2015, revaz+jablonka2018}. However,
such processes would not explain why stellar halo formation may be
non-universal. If a stellar halo was formed in an older, induced burst
of star formation, then the star formation history should contain
evidence of a corresponding burst.

\textbf{Major mergers.} Dwarf galaxy mergers may be relatively common
\citep{deason+wetzel+garrison-kimmel2014}. After a merger, stars from
the lower-mass galaxy are preferentially dispersed, forming a stellar
halo and population gradient \citep{benitez-llambay+2016, deason+2022}.
If the galaxies contains gas, the merger can trigger new star formation,
forming a younger population of stars \citep[e.g.,][]{genina+2019}. A
few local dwarfs are suspected to have undergone a major merger,
including Tucana II, Andromeda II, and Phoenix
\citep{lokas+2014, fouquet+2017, tarumi+2021, cardona-barrero+2021, querci+2025}.
A galaxy having undergone a major merger should show evidence of at
least two or three populations from distinct origins.

\textbf{Minor mergers / accretion}. Just as the Milky Way's halo is
believed to be built from many minor mergers, dwarf galaxies may build
stellar halos composed of even fainter satellites
\citep{ricotti+polisensky+cleland2022}. Few examples are known, but a
stream around And II may be an instance of dwarf accretion
\citep{amorisco+evans+vandeven2014, roederer+2023}. In this scenario,
the stellar halo's would contain chemical signatures from several
distinct ultra-faint dwarf galaxies.

@@

In both cases, tides more plausibly produce extended density profiles.
However, recovering the observed velocity dispersions often requires
ongoing tidal disruption
\citetext{\citealp{mcgaugh+wolf2010}; \citealp[but see
also][]{sanchez-salcedo+hernandez2007}}. Current data do not show
evidence of such features (see Section~\ref{sec:extra_rv_models}), so
future data would need to uncover signs of disruption to support this
hypothesis.

\textbf{Dynamical heating.} Old stars in dwarf galaxies may be hotter
than younger stars due to processes including stellar feedback
\citep{stinson+2009, maxwell+2012, el-badry+2016, mercado+2021},
sub-subhalo interactions \citep{penarrubia+2025}, or even fuzzy dark
matter interference fringes
\citep[e.g.,][]{el-zant+2020, duttachowdhury+2023}. However, most of
these processes should operate similarly across dwarf galaxies---stellar
halos should be more common if these processes form substantial halos.

\section{Disentangling the origin of a stellar
halo}\label{disentangling-the-origin-of-a-stellar-halo}

We just reviewed a number of different, possibly-concurrent explanations
for the formation of a stellar halo. While we leave the nature of Scl
and UMi's extended stellar densities an open question, we can discuss
possible clues to different formation scenarios.

@@

Ongoing and future facilities will be essential for testing dwarf galaxy
halo formation. Given the typical faint magnitudes of dSph stars,
chemistry is best accessible through multi-object spectroscopic
instruments on large telescopes (e.g., 4MOST and extremely large
telescopes \citet{skuladottir+2023}). 3D internal kinematics of dwarfs
may require a successor to \emph{Gaia}. However, using JWST and HST
enables precise proper motions for small regions within dwarf galaxies
\citep[e.g.,][]{vitral+2025}, which may be sufficient to constrain
vastly different internal dynamic structures and ongoing tidal
disruption. Finally, ongoing photometric surveys by Rubin Observatory
\citep{ivezic+2019} and the Euclid mission
\citep{euclidcollaboration+2025} will probe even fainter features around
dwarf galaxies, possibly finding more halos, tidal streams, and
improving our understanding of their stellar populations. Altogether,
new surveys will likely uncover novel aspects about the inner workings
and outer halos of dwarf galaxies.

The next generation of dwarf simulations are beginning to answer
questions and unravel processes in the formation of dwarf galaxies.
Notable recent simulation projects include EDGE \citep{agertz+2020} and
LYRA \citep{gutcke+2021}, both of which can simulate the cosmological
formation of isolated dwarf galaxies while resolving individual
(massive) stars. By including more realism (e.g., black holes, magnetic
fields), refining models of poorly understood processes (feedback and
star formation), and improving numerical resolution, simulations will
continue to explore these processes and refine our understanding of
dwarf galaxy formation.

@@

Finally, we discuss the implications of a non-tidal density excess in
Scl and UMi. We review mechanisms ranging from episodic star formation
histories, to accretion, to departures from \LCDM{}. While a more
precise explanation awaits, upcoming and ongoing surveys will uncover
the stellar populations within dwarf galaxies in unprecedented detail.
Future work will continue to uncover the cosmic origins of the smallest
structures forming building blocks of galaxies like ours.

>>>>

In the previous chapters, we have shown that Sculptor and Ursa Minor
have more extended density profiles compared to other MW classical
satellites, and that recent tides are an unlikely explanation for such
extended density profiles. We therefore suggest that Scl and UMi may
host an additional, extended population of stars. With our data, other
classical dwarfs lack evidence for similar extended populations. The
absence of similar extended populations in other classical dwarfs
suggests Scl and UMi have followed a divergent evolutionary path.

In this chapter, we place our results in context with past work. We
review both work considering tidal effects on Scl and UMi and detections
of peculiarities. We then consider possible origins of extended stellar
populations for these galaxies. Scenarios include multi-epoch star
formation, past mergers depositing stars into the outskirts, and even
departures from the typical exponential dwarf galaxy structure. We later
discuss how future observations and simulations may distinguish halo
formation scenarios. Finally, we summarize the main conclusions of this
thesis.

@@

Tidal effects have been long invoked to explain features of dwarf
galaxies e.g.~Section \ref{sec:scl_umi_obs_tides},
\citep{mayer+2001a, tsujimoto+shigeyama2002} \textbf{Check this
citation}. Yet, growing work suggests tides may not be as commonplace as
suspected. For example, \citet{read+2006} suggest that the absence of
rising radial velocity dispersion profiles in dwarfs implies a lack of
tidal disruption. \citet{penarrubia+2009}, introducing the break radius
as a diagnostic (Eq.~\ref{eq:r_break}), showed that most satellite
orbits were inconsistent with observable tidal effects. Recently,
\citet{pace+erkal+li2022} use a Jacobi radius-based criterion
(Eq.~\ref{eq:r_jacobi}) to reach a similar conclusion. With the
exception of the Sagittarius dSph, these results suggest tides have mild
influence on the stellar components of classical dwarfs.

@@

Our work extends these models, adding updated structural properties, a
broader range of orbital histories, and the influence of the LMC.
Despite these considerations, we still reach similar conclusions: tidal
effects do not shape Scl or UMi's stellar component. A non-tidal
explanation likely underlies the origin of these galaxies' outer density
excess.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_umi_fe_h_gradient.pdf}}
\caption[Metallicity gradients in Scl and UMi]{Density profiles and
metallicity gradients in Scl (left) and UMi (right). In the top panel,
we plot the J+24 density profiles as function of elliptical radius in
black, and the two components of a double-exponential density fit as
curved, transparent lines. The vertical lines represent the transition
from the inner to the outer exponential profile. \textbf{Bottom}:
Metallicities as a function of elliptical radius for stars from
\citet{tolstoy+2023}; \citet{abdurrouf+2022}; and \citet{sestito+2023a}
(left) and \citet{pace+2020}; \citet{abdurrouf+2022}; and
\citet{sestito+2023b} (right), crossmatched against J+24 as described in
Appendix \ref{sec:extra_rv_models}.}\label{fig:metallicity_gradients}
\end{figure}

@@

Could these internal stellar populations map to the observed density
excess? In Scl, both populations have scale radii likely smaller than
the half-light radius. But, the third, more tentatively detected
population in \citet{arroyo-polonio+2024} would reside within the outer
density excess. In Ursa Minor, the transition between the metal-rich and
metal-poor stellar components occurs at \(\sim 30\,\)arcmin
\citep{pace+2020}, which coincides with the density excess. This
metal-poor population may thus be related to the extended stellar
density of Ursa Minor.

Fig.~\ref{fig:metallicity_gradients} shows the radial metallicity
distribution in both galaxies. To aid in comparison, we fit a
2-component exponential to the density profiles. We also mark where the
densities of each component become equal, the ``transition'' radius. Scl
and UMi both show evidence of a metallicity gradient. But, the relation
between the metallicity gradient and the transition from the inner to
the outer density profile is unclear. UMi also has few measurements past
the transition radius. Further observations would be necessary to
understand if the extended stellar component represents a chemically
distinct population.

Ursa Minor has also shown evidence for possible inner substructure such
as stellar or kinematic ``clumps''
\citep[e.g.,][]{olszewski+aaronson1985, demers+1995, kleyna+1998, battinelli+demers1999, bellazzini+2002}.
While one clump was re-detected kinematically in \citet{pace+2014},
\citet{munoz+2018} find no evidence of substructure with modern
photometry. In any case, the survival of a cold clump depends on the
inner structure of the halo, disrupting quickly in halos with
substructure or inner density cusps but survining longer in smooth,
cored halos \citep{kleyna+2003, lora+2012}.

Given their multiple stellar populations and possible inner
substructure, this hints that Scl and UMi's distinct structure today
arises from an unusual past. For example, a merger may naturally explain
both extended and multiple stellar populations.

\section{Forming an extended stellar
population}\label{sec:stellar_halos}

As we disfavour a tidal origin for Scl and UMi's extended density
profiles, we now consider six alternative processes which may form an
extended stellar population in these dwarfs.

\textbf{Episodic star formation.} Star formation may quench and
reignite, creating successive stellar generations with differing
distributions. External star formation triggers include tidal
compression \citep{mayer+2001a, dong+lin+murray2003}, collisions with
gaseous filaments \citep{genina+2019}, perturbations from dark halos
\citep{starkenburg+helmi+sales2016}, or shocks with the MW corona
\citep{wright+2019}. More common mechanisms, like feedback or
reionization-driven quenching, may also form multiple stellar
generations
\citep{kawata+2006, benitez-llambay+2015, revaz+jablonka2018}. However,
such processes would not explain why extended stellar populations appear
to be non-universal. If an extended population was formed in an older,
induced burst of star formation, then the star formation history should
contain evidence of a corresponding burst.

\textbf{Major mergers.} Dwarf galaxy mergers may be relatively common
\citep{deason+wetzel+garrison-kimmel2014}. After a merger, stars from
the lower-mass galaxy are preferentially dispersed, forming an extended
stellar component and population gradient
\citep{benitez-llambay+2016, deason+2022}. If the galaxies contains gas,
the merger can trigger new star formation, forming a younger population
of stars \citep[e.g.,][]{genina+2019}. A few local dwarfs are suspected
to have undergone a major merger, including Tucana II, Andromeda II, and
Phoenix
\citep{lokas+2014, fouquet+2017, tarumi+yoshida+frebel2021, cardona-barrero+2021, querci+2025}.
A galaxy having undergone a major merger should show evidence of at
least two or three populations from distinct origins.

Just as the Milky Way's halo is believed to be built from many minor
mergers, dwarf galaxies may build accreted ``stellar halos'' composed of
even fainter satellites \citep{ricotti+polisensky+cleland2022}. Few
examples are known, but a stream around And II may be an instance of
dwarf accretion \citep{amorisco+evans+vandeven2014, roederer+2023}. In
this scenario, the extended stellar populations would contain chemical
signatures from several distinct ultra-faint dwarf galaxies.

@@

In both cases, tides more plausibly produce extended density profiles.
However, recovering the observed velocity dispersions often requires
ongoing tidal disruption
\citetext{\citealp{mcgaugh+wolf2010}; \citealp[but see
also][]{sanchez-salcedo+hernandez2007}}. Current data do not show
evidence of such features (see Appendix \ref{sec:extra_rv_models}), so
future data would need to uncover signs of disruption to support this
hypothesis.

\textbf{Dynamical heating.} Old stars in dwarf galaxies may be hotter
than younger stars due to processes including stellar feedback
\citep{stinson+2009, maxwell+2012, el-badry+2016, mercado+2021},
sub-subhalo interactions \citep{penarrubia+2025}, or even fuzzy dark
matter interference fringes
\citep[e.g.,][]{el-zant+2020, duttachowdhury+2023}. However, most of
these processes should operate similarly across dwarf
galaxies---extended stellar populations should be more common if these
processes form substantial halos.

\section{Disentangling the origin of extended stellar
populations}\label{disentangling-the-origin-of-extended-stellar-populations}

We reviewed a number of different, possibly-concurrent explanations for
the formation of extended populations. While we leave the nature of Scl
and UMi's extended stellar densities an open question, we can discuss
possible clues to different formation scenarios.

@@

Ongoing and future observational surveys will be instrumental for
testing dwarf galaxy halo formation. Because dSph stars are usually
faint, detailed chemical analysis will rely on medium--high resolution
multi-object spectroscopy on large telescopes, such as 4MOST
\citep{skuladottir+2023} or extremely large telescopes. Mapping the 3D
kinematics across a dwarf may require a \emph{Gaia} successor, but
ongoing surveys with JWST and HST can derive precise proper motions
within select regions of dwarf galaxies, able to reconstruct internal
structure and anisotropies \citep[e.g.,][]{vitral+2025}. Current
wide-field photometric surveys by Rubin Observatory \citep{ivezic+2019}
and the Euclid mission \citep{euclidcollaboration+2025} will trace even
fainter features around dwarf galaxies, possibly finding halos, tidal
streams, and unusual stellar populations. Together, these observational
programs promise a new window into the detailed structure, evolution,
and outskirts of dwarf galaxies.

The next generation of dwarf simulations will likewise continue to
refine our understanding of dwarf galaxy formation. Notable recent
simulation projects include EDGE \citep{agertz+2020} and LYRA
\citep{gutcke+2021}, both of which can simulate the cosmological
formation of isolated dwarf galaxies while resolving individual
(massive) stars. Next steps for simulations include improved realism
(e.g., magnetic fields), refining physical models (including feedback
and star formation), and improving numerical resolution. With these
continuing advances, future simulations may shed light on viability of
halo formation scenarios and the details of dwarf galaxy assembly.

@@

Finally, we discuss the implications of a non-tidal density excess in
Scl and UMi. We review mechanisms ranging from episodic star formation
histories, to accretion, to departures from \LCDM{}. Future simulations
and observational programs will be able to investigate the nature of
dwarfs with ever increasing detail. Abundant opportunities await to
uncover the cosmic origins and history of dwarf galaxies, the building
blocks of galaxies like ours.
